---
title: "Examples"
output: html_document
date: "2024-07-04"
---
# SeroMetrics

## I. Introduction

### Data Format
#### - participant id: mandatory. This should be the id of the sampled subject.
#### - value: mandatory. This should be the investigated value, usually being titers.
#### - distance: mandatory. This should be the absolute distance among virus strains, usually being isolation year or AA distance.
#### - strain: optional. This should be the virus strains. When distance is not provided, one may need to derive isolation year from the information in virus strain.
#### - group: optional. This should be the group of some sample of a participant. Sometimes, one participant may take tests on the same strains multiple times, and the samples collected from each time forms a sample group.
#### - other variables: optional. There might be age, sex, and other information regarding one sample. At most times, these variables will not affect the final result of the metrics.

### Data Simulation (Sample codes)
```{r}
library(dplyr)

set.seed(12986)

id <- rep(1:30, each = 2)

unique_birth_year <- sample(1937:2014, size = 30, replace = TRUE)

birth_year <- rep(unique_birth_year, each = 2)

sample_year <- rep(c(2020, 2024), times = 30)

df <- data.frame(id = id, birth_year = birth_year, sample_year = sample_year)

unique_isolation_years <- sample(1950:2020, size = 20, replace = TRUE)

df <- df %>%
  slice(rep(1:n(), each = 20)) %>%
  mutate(isolation_year = rep(unique_isolation_years, times = n() / 20))

generate_log_titer <- function(birth_year, isolation_year) {
  if(isolation_year<1960){
    if(birth_year > 2005){
      return(sample(-1:2,1))
    }else if(birth_year >=1995 & birth_year < 2005){
      return(sample(2:5,1))
    }else if(birth_year >=1955 & birth_year < 1995){
      return(sample(5:6,1))
    }else{
      return(sample(4:6,1))
    }
  }else if(isolation_year >= 1960 & isolation_year<=1970){
    if(birth_year > 2005){
      return(sample(-1:2,1))
    }else if(birth_year >=1995 & birth_year < 2005){
      return(sample(3:5,1))
    }else if(birth_year >=1955 & birth_year < 1995){
      return(sample(6:7,1))
    }else{
      return(sample(5:6,1))
    }
  }else if(isolation_year >= 1970 & isolation_year<=1995){
    if(birth_year > 2005){
      return(sample(-1:2,1))
    }else if(birth_year >=1995 & birth_year < 2005){
      return(sample(2:6,1))
    }else if(birth_year >=1955 & birth_year < 1995){
      return(sample(4:5,1))
    }else{
      return(sample(3:4,1))
    }
  }else if(isolation_year >= 1995 & isolation_year<=2005){
    if(birth_year > 2005){
      return(sample(0:4,1))
    }else if(birth_year >=1995 & birth_year < 2005){
      return(sample(6:7,1))
    }else if(birth_year >=1955 & birth_year < 1995){
      return(sample(4:7,1))
    }else{
      return(sample(3:5,1))
    }
  }else if(isolation_year >= 2005){
    if(birth_year > 2005){
      if(isolation_year > birth_year){
        return(sample(5:8,1))
      }else{
        return(sample(2:6,1))
      }
    }else if(birth_year >=1995 & birth_year < 2005){
      return(sample(7:8,1))
    }else if(birth_year >=1955 & birth_year < 1995){
      return(sample(2:5,1))
    }else{
      return(sample(0:4,1))
    }
  }
}

df <- df %>%
  rowwise() %>%
  mutate(log_titer = generate_log_titer(birth_year, isolation_year)) %>%
  ungroup()

head(df,10)
```

## II. Environment Build-up

### Step 1: Clear up the global environment.

### Step 2: Download the R folder and store it in a folder called SeroMetrics.

### Step 2: Run the codes below.
```{r}
# Download necessary packages if you have not installed them in your PC
# install.packages("roxygen2)
# install.packages("devtools")
# install.packages("usethis")

library(roxygen2)
library(devtools)
library(usethis)

create_package("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics")
roxygenise("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics")
build("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics")
install("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics")

library(SeroMetrics)
```

## III. Examples of real datasets
### 1. Lessler
#### This is a basic example with a simple dataset, in which for each participant, there is only one testing result on each virus strain, i.e."group_col" can be NULL, and all the virus strains have a different isolation year, i.e. "mode" can be NULL.

##### The first method is to use the overallData function directly.
```{r}
library(readxl)
library(stringr)
library(DescTools)
library(moments)
library(dplyr)

#load data
df <- read_excel("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/Lessler.xlsx")
#data pre-processing
df$isolation_year <- as.numeric(str_extract(df$neut.against, "\\d{4}"))
df$titers <- floor(df$titers)
#show the basic structure of the dataset
head(df,10)
```

###### You can choose the metrics as you want. However, there are some restrictions when inputting the arguments. For details, check the explanation of the function overallData.
```{r}
overallResult <- overallData(df, part_col = "id", weight_col = "isolation_year", val_col = "titers", var_trans = 1, gmt_output_trans = 1, required_metrics = c("AUC","ATY","GMT","Prot_prop","Width0","Width2","Width3"), col_names = c("auc","aty","gmt","prot_prop","Width10","Width40","Width80"))
head(overallResult, 10)
```

##### The second method is to calculate the metrics one by one, and join them manually.

```{r}
library(readxl)
library(stringr)
library(DescTools)
library(moments)
library(dplyr)

#load data
df <- read_excel("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/Lessler.xlsx")
#data pre-processing
df$isolation_year <- as.numeric(str_extract(df$neut.against, "\\d{4}"))
df$titers <- floor(df$titers)
#show the basic structure of the dataset
head(df,10)
```
###### Choose the metrics as you want.
```{r}
atyResult <- atyData(df,part_col = c("id"),weight_col = c("isolation_year"), val_col = c("titers"), var_trans = 1)
aucResult <- aucData(df,part_col = c("id"),weight_col = c("isolation_year"), val_col = c("titers"), var_trans = 1)
giniResult <- giniData(df,part_col = c("id"), val_col = c("titers"), var_trans = 1, gini_col = "Gini")
gmtResult <- gmtData(df,part_col = c("id"), val_col = c("titers"), var_trans = 1,output_trans = 1)
kurtosisResult <- kurtosisData(df,part_col = c("id"), weight_col = c("isolation_year"), val_col = c("titers"), var_trans = 1)
max_titerResult <- max_titerData(df,part_col = c("id"), val_col = c("titers"), var_trans = 1,output_trans = 1)
propResult2 <- propData(df,part_col = c("id"), val_col = c("titers"), var_trans = 1,threshold_trans = 1,threshold = 2, prop_col = "Prop2") #the proportion of titers above 2
propResult3 <- propData(df,part_col = c("id"), val_col = c("titers"), var_trans = 1,threshold_trans = 1,threshold = 3, prop_col = "Prop3") #the proportion of titers above 3
prot_propResult <- prot_propData(df,part_col = c("id"), val_col = c("titers")) #using default weight vector
skewnessResult <- skewnessData(df,part_col = c("id"), weight_col = c("isolation_year"), val_col = c("titers"), var_trans = 1)
widthResult <- widthData(df,part_col = c("id"), weight_col = c("isolation_year"), val_col = c("titers"), var_trans = 1, threshold_trans = 1,threshold = 2, width_col = "Width2") #the width of titers above 2

overallResult <- atyResult %>%
  left_join(aucResult, by = "id") %>%
  left_join(giniResult, by = "id") %>%
  left_join(gmtResult, by = "id") %>%
  left_join(kurtosisResult, by = "id") %>%
  left_join(max_titerResult, by = "id") %>%
  left_join(propResult2, by = "id") %>%
  left_join(propResult3, by = "id") %>%
  left_join(prot_propResult, by = "id") %>%
  left_join(skewnessResult, by = "id") %>%
  left_join(widthResult, by = "id")

head(overallResult,10)
```

### 2. Fonville
#### This is a more complicated example with the dataset called "Fonville", in which for each participant, there are multiple testing results, one for each sample year, i.e. "group_col" is sample year, and there are different virus strains share the same isolation_year, i.e. "mode" is necessary. Besides, the dataset has a particular structure with each virus strain forming one column, so data transformation should be done first.

##### The first method is to use the overallData function directly.
```{r}
library(readxl)
library(stringr)
library(DescTools)
library(moments)
library(dplyr)
library(tidyverse) #this is for data transformation

#This transformation R-script specifically works for the Fonville dataset. If one's dataset has a structure different than the standard one, one may need to write the transformation codes by oneself, and use the package after transforming the data.
source('C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/transform.R')

#load valid data
df <- read_excel("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/Fonville.xlsx", n_max = 324)
#data transformation
transformed_df <- transform(df)

head(transformed_df,10)
```

###### Choose the metrics as you want.
```{r}
overallResult <- overallData(transformed_df, part_col = "Subject Number", weight_col = "isolation_year", val_col = "titer", group_col = "Sample Year", var_trans = 0, max_titer_output_trans = 1, mode = "mean", required_metrics = c("gini_coefficient","Kurtosis","Skewness","Prop0","Prop2","Prop3"), col_names = c("Gini","Kurt","Skew","prop0","prop2","prop3"))
head(overallResult, 10)
```

##### The second method is to calculate the metrics one by one, and join them manually.
```{r}
library(readxl)
library(stringr)
library(DescTools)
library(moments)
library(dplyr)
library(tidyverse) #this is for data transformation

#This transformation R-script specifically works for the Fonville dataset. If one's dataset has a structure different than the standard one, one may need to write the transformation codes by oneself, and use the package after transforming the data.
source('C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/transform.R')

#load valid data
df <- read_excel("C:/Users/xShirleyx/Desktop/SRA/20240717/SeroMetrics/data/Fonville.xlsx", n_max = 324)
#data transformation
transformed_df <- transform(df)

head(transformed_df,10)
```
###### Choose the metrics as you want.
```{r}
atyResult <- atyData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0, aty_col = "aty")
aucResult <- aucData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0,mode="mean", auc_col = "auc")
giniResult <- giniData(transformed_df,part_col = c("Subject Number"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0)
gmtResult <- gmtData(transformed_df,part_col = c("Subject Number"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0,output_trans = 1)
kurtosisResult <- kurtosisData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0)
max_titerResult <-max_titerData(transformed_df,part_col = c("Subject Number"), val_col = c("titer"),group_col = c("Sample Year"),var_trans = 0,output_trans = 1)
propResult <- propData(transformed_df,part_col = c("Subject Number"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0, threshold_trans = 1,threshold = 2) #the proportion of titers above 2
prot_propResult <- prot_propData(transformed_df,part_col = c("Subject Number"),val_col = c("titer"),group_col = c("Sample Year"), min.y = 0, weight = c(0,0.3,0.5,0.8,0.9))
skewnessResult <- skewnessData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0)
widthResult2 <- widthData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0,threshold_trans = 1,threshold = 2,mode="mean",width_col = "Width2") #the width of titers above 2
widthResult3 <- widthData(transformed_df,part_col = c("Subject Number"),weight_col = c("isolation_year"), val_col = c("titer"), group_col = c("Sample Year"),var_trans = 0,threshold_trans = 1,threshold = 3,mode="mean",width_col = "Width3") #the width of titers above 3

overallResult <- atyResult %>%
  left_join(aucResult, by = c("Subject Number","Sample Year")) %>%
  left_join(giniResult, by = c("Subject Number","Sample Year")) %>%
  left_join(gmtResult, by = c("Subject Number","Sample Year")) %>%
  left_join(kurtosisResult, by = c("Subject Number","Sample Year")) %>%
  left_join(max_titerResult, by = c("Subject Number","Sample Year")) %>%
  left_join(propResult, by = c("Subject Number","Sample Year")) %>%
  left_join(prot_propResult, by = c("Subject Number","Sample Year")) %>%
  left_join(skewnessResult, by = c("Subject Number","Sample Year")) %>%
  left_join(widthResult2, by = c("Subject Number","Sample Year"))%>%
  left_join(widthResult3, by = c("Subject Number","Sample Year"))

head(overallResult,10)
```
