---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# SeroMetrics

`SeroMetrics` quantifies an individual's *antibody landscape* — the profile of
haemagglutination inhibition (HAI) titers measured against a panel of influenza
viruses isolated at different times or in different antigenic spaces. It
provides a consistent set of summary metrics that capture the magnitude,
breadth, skewness, and tailedness of an antibody response, computed directly
from titer data.

Summary metrics and application examples are adapted from:

-   Yang B, Lessler J, Zhu H, Jiang CQ, Read JM, Hay JA, Kwok KO, Shen R, Guan Y,
    Riley S, Cummings DA. [Life course exposures continually shape antibody
    profiles and risk of seroconversion to
    influenza](https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1008635).
    PLoS Pathogens. 2020;16(7):e1008635.

-   Yang B, Leung NHL, Fox A, Sullivan SG, Ho F, Barr IG, Cheng SMS, Wong SS,
    Levine MZ, Iuliano AD, Peiris M, Thompson MG, Cowling BJ. Evaluating
    Antibody Breadth Following Various Enhanced Influenza Vaccines in Older
    Adults. *In preparation*.

## Installation

Install the released version from CRAN:

```r
install.packages("SeroMetrics")
```

Or the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("byyangyby/SeroMetrics")
```

## Available metrics

| Metric | Function | Captures |
|---|---|---|
| GMT | `gmt_data()` | Magnitude (geometric mean) |
| Max titer | `max_titer_data()` | Peak response |
| AUC | `auc_data()` | Breadth × magnitude |
| ATY | `aty_data()` | Average response year (temporal centre) |
| Width | `width_data()` | Span of strains above a threshold |
| Proportion | `prop_data()` | Fraction of strains above a threshold |
| Protected proportion | `prot_prop_data()` | Fraction weighted by protection levels |
| Gini coefficient | `gini_data()` | Inequality of the titer distribution |
| Skewness | `skewness_data()` | Asymmetry of the landscape |
| Kurtosis | `kurtosis_data()` | Peakedness / tailedness |

`summarise_landscape()` computes several metrics at once.

## Data structure

Each function expects a **long-format** data frame — one row per
participant–strain pair — identified through these arguments:

| Argument | Required | Description |
|:---|:---|:---|
| `part_col` | yes | Participant identifier. |
| `val_col` | yes | The measured value, usually a titer. |
| `weight_col` | for AUC, ATY, Width, Skewness, Kurtosis | Strain distance — e.g. isolation year, or amino-acid distance to a reference strain. |
| `group_col` | optional | Sample group, used when a participant has multiple serum samples (e.g. sample year). |
| other | optional | Age, sex, and other covariates; ignored by the metric functions. |

## Bundled datasets

The package ships two real HAI titer datasets, already in long format:

```{r datasets}
library(SeroMetrics)

data(Fonville)
head(Fonville[, c("Subject Number", "Sample Year", "isolation_year",
                  "strain", "titer")])

data(Lessler)
head(Lessler[, c("id", "isolation_year", "strain", "titer")])
```

## Example 1 — Fonville et al. 2014, *Science*

[Fonville et al., 2014](https://www.science.org/doi/10.1126/science.1256427)
collected longitudinal antisera over six years, so each participant has several
serum samples. Set `group_col = "Sample Year"`.

### Calculate multiple metrics at once

`summarise_landscape()` returns one row per participant–sample, with a column for each
requested metric.

```{r fonville-overall}
summarise_landscape(
  Fonville,
  part_col         = "Subject Number",
  weight_col       = "isolation_year",
  val_col          = "titer",
  group_col        = "Sample Year",
  var_trans        = TRUE,
  required_metrics = c("GMT", "AUC", "ATY", "gini_coefficient",
                       "Kurtosis", "Skewness", "Prop2", "Width2")
) |>
  head()
```

### Calculate an individual metric

Each metric also has its own function. They share the same column arguments and
can be joined on the participant (and group) columns.

```{r fonville-individual}
gmt_res  <- gmt_data(Fonville,  part_col = "Subject Number", val_col = "titer",
                    group_col = "Sample Year")
auc_res  <- auc_data(Fonville,  part_col = "Subject Number",
                    weight_col = "isolation_year", val_col = "titer",
                    group_col = "Sample Year")

head(gmt_res)
head(auc_res)
```

## Example 2 — Lessler et al. 2012, *PLOS Pathogens*

[Lessler et al., 2012](https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1002802)
collected one serum sample per participant against a panel of 9 H3N2 strains
spanning 1968–2008, so no `group_col` is needed. Titers (`titer`) are already
log-transformed, so leave `var_trans = TRUE`.

### Calculate multiple metrics at once

```{r lessler-overall}
summarise_landscape(
  Lessler,
  part_col         = "id",
  weight_col       = "isolation_year",
  val_col          = "titer",
  var_trans        = TRUE,
  required_metrics = c("GMT", "AUC", "gini_coefficient",
                       "Kurtosis", "Skewness", "Prop2", "Width2")
) |>
  head()
```

### Calculate an individual metric

```{r lessler-individual}
auc_data(Lessler, part_col = "id", weight_col = "isolation_year",
        val_col = "titer") |>
  head()
```

### Aggregating duplicate isolation years with `mode`

If your dataset has multiple titers against different strains that share an
isolation year (as in `Fonville`, where most participant-year combinations
have several strains), set `mode = "mean"`, `"max"`, or `"min"` to aggregate
them before computing distance-weighted metrics (AUC, ATY, Width, Skewness,
Kurtosis):

```{r fonville-mode-mean}
summarise_landscape(
  Fonville,
  part_col         = "Subject Number",
  weight_col       = "isolation_year",
  val_col          = "titer",
  group_col        = "Sample Year",
  mode             = "mean",
  required_metrics = c("AUC", "ATY", "Skewness", "Kurtosis")
) |>
  head()
```

## Learn more

See the introductory vignette for a full walkthrough of every metric and the
key parameters (`var_trans`, `base`, `adj`, `mode`, `adjust`):

```r
vignette("SeroMetrics")
```
