---
title: "Introduction to SeroMetrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SeroMetrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SeroMetrics)
```

## Overview

SeroMetrics quantifies antibody landscapes — the profile of antibody titers a
person has against a panel of virus strains — using a set of summary metrics.
Each metric captures a different aspect of the landscape shape:

| Metric | Function | Captures |
|---|---|---|
| GMT | `gmt_data()` | Magnitude (geometric mean) |
| Max titer | `max_titer_data()` | Peak response |
| AUC | `auc_data()` | Breadth × magnitude |
| ATY | `aty_data()` | Average response year (temporal center) |
| Width | `width_data()` | Span of strains above a threshold |
| Proportion | `prop_data()` | Fraction of strains above a threshold |
| Protected proportion | `prot_prop_data()` | Fraction weighted by protection levels |
| Gini coefficient | `gini_data()` | Inequality of titer distribution |
| Skewness | `skewness_data()` | Asymmetry of the landscape |
| Kurtosis | `kurtosis_data()` | Peakedness / tailedness |

---

## Data Format

Each function expects a **long-format** data frame with at minimum:

- **`part_col`** — participant ID (one row per participant–strain pair)
- **`val_col`** — titer value (log-transformed by default; set `var_trans = FALSE` for raw titers)
- **`weight_col`** — strain distance (e.g. isolation year) — required for AUC, ATY, Width, Skewness, Kurtosis
- **`group_col`** — sample group (e.g. sample year) — use when a participant has multiple serum samples

---

## Example 1 — Fonville Dataset (longitudinal)

The `Fonville` dataset contains longitudinal HI titers. Participants were sampled
across multiple years, so use `group_col = "Sample Year"`.

```{r fonville-load}
data(Fonville)
head(Fonville[, c("Subject Number", "Sample Year", "isolation_year", "strain", "titer")], 6)
```

### Calculate all metrics at once with `summarise_landscape()`

```{r fonville-overall}
result <- summarise_landscape(
  Fonville,
  part_col        = "Subject Number",
  weight_col      = "isolation_year",
  val_col         = "titer",
  group_col       = "Sample Year",
  var_trans       = TRUE,
  required_metrics = c("GMT", "AUC", "ATY", "gini_coefficient",
                       "Kurtosis", "Skewness", "Prop2", "Width2")
)
head(result, 6)
```

### Calculate metrics individually

```{r fonville-individual}
gmt_res  <- gmt_data(Fonville,  part_col = "Subject Number", val_col = "titer",
                    group_col = "Sample Year")
auc_res  <- auc_data(Fonville,  part_col = "Subject Number",
                    weight_col = "isolation_year", val_col = "titer",
                    group_col = "Sample Year")
gini_res <- gini_data(Fonville, part_col = "Subject Number", val_col = "titer",
                     group_col = "Sample Year")

head(gmt_res, 4)
head(auc_res, 4)
```

---

## Example 2 — Lessler Dataset (duplicate strain years)

In the `Lessler` dataset each participant (`id`) has titers against a panel
of strains, and several strains share the same isolation year. Use
`mode = "mean"` to aggregate titers with duplicate isolation years before
computing distance-weighted metrics.

```{r lessler-load}
data(Lessler)
head(Lessler[, c("id", "isolation_year", "strain", "titer")], 6)
```

```{r lessler-overall}
result2 <- summarise_landscape(
  Lessler,
  part_col        = "id",
  weight_col      = "isolation_year",
  val_col         = "titer",
  var_trans       = TRUE,
  mode            = "mean",
  required_metrics = c("GMT", "AUC", "gini_coefficient",
                       "Kurtosis", "Skewness", "Prop2", "Width2")
)
head(result2, 6)
```

---

## Core metric functions

Each `*Data()` function wraps a lower-level scalar function of the same root
name that operates directly on vectors. These are useful for custom workflows:

```{r core-functions}
titer <- c(3, 5, 6, 4, 2)   # log2-transformed titers
year  <- c(1968, 1972, 1977, 1987, 1997)

gmt(titer)
auc(titer, year, input.log.trans = TRUE)
aty(titer, year, input.log.trans = TRUE)
prop(titer, threshold = 4)
```

---

## Key parameters

| Parameter | Default | Meaning |
|---|---|---|
| `var_trans` | `TRUE` | Input titers are already log-transformed |
| `output_trans` | `TRUE` | Return log-transformed output (GMT, Max titer) |
| `base` | `2` | Log base for transformation |
| `adj` | `10` | Divisor before log transform (raw titer → log2(titer/10)) |
| `mode` | `NULL` | How to resolve duplicate `weight_col` values: `"mean"`, `"max"`, `"min"` |
| `adjust` | `NULL` | Titer shift before area calculations; defaults to min titer |
